Modelling the Lymphatic Metastatic Progression Pathways of OPSCC \\ from Multi-Institutional Datasets

### 2023-12-20 src/tex/ms.tex
[showyourwork] overleaf sync

@@

\begin{figure}
    \script{thermo_int.py}
    \centering
    \includegraphics[width=\linewidth]{figures/thermo_int.png}
    \caption{The expected log-accuracy under the power posterior plotted against the value of the inverse temperature $\beta$. These values were computed during a \gls{ti} run for the loosing graph model (red line), as well as the winning graph model (green line). Note that the scaling of the x-axis is chosen such that the 64 points on the temperature ladder appear evenly spaced. This is to stress how the log-accuracy develops in the range from $\beta=0$ to around $\beta=0.1$. The area under these two curves yields an approximation of the log-evidence of the respective model. This plot one indicates that due to the simplicity of the loosing graph model its accuracy increases earlier and faster that the winning graph model's. However, already before $\beta=0.031$ the model based on the winning graph outperforms the loosing graph model. Ultimately, the winning graph's log-evidence is better than the loosing graph model's by a value of around \variable{output/evidence_diff.txt}(see \autoref{table:evidence}). \label{fig:thermo_int}}
\end{figure}

@@

@@

\begin{deluxetable}{ccccccrrrr}
    \tablecaption{Prevalence of involvement patterns in the multi-centric dataset. An involvement pattern is characterized by the state of the six \glspl{lnl}: A red dot means the \gls{lnl} was reported to be metastatic, a green dot means it was determined to be healthy and a question mark means that the prevalence was marginalized over the state of this \gls{lnl}. \label{table:data_prevalence}}
    \tablehead{
        \multicolumn{6}{c}{\textbf{LNL involvement} } & \multicolumn{4}{c}{\textbf{T-category} } \\
        I & II & III & IV & V & VII & \multicolumn{2}{c}{\textbf{early} } & \multicolumn{2}{c}{\textbf{advanced} }
    }
    \startdata
    \variable{output/data_table.tex}
    \enddata
\end{deluxetable}

@@

The dataset containing the consensus decision for the involvement of each level in every patient was then used for model parameter learning. We assumed that it represents an observation of the true hidden state $\boldsymbol{\xi}_k$, which corresponds to setting the sensitivity and specificity both to one. The frequencies of some of the most important combinations of involved lymph node levels are listed in \autoref{table:data_prevalence}.

@@

\begin{figure*}
    \script{bg_core_prevs.py}
    \begin{centering}
        \includegraphics[width=\textwidth]{figures/bg_core_prevs.png}
        \caption{Prevalence of involvement as predicted by the base graph model for different scenarios involving the most commonly metastatic \glspl{lnl} II, III and IV (shaded histograms). The model's predictions are compared to Beta posteriors over the prevalence based on the frequency of the same scenarios and a uniform prior (slid lines). The top panel shows some selected scenarios with early T-category tumors and the bottom panel with advanced T-category. \label{fig:bg_prevalences}}
    \end{centering}
\end{figure*}

@@

@@

\begin{figure*}
    \centering
    \includegraphics[width=\linewidth]{figures/ranking.pdf}
    \caption{Visual ranking of the investigated graphs w.r.t. their model evidence, computed via thermodynamic integration. Not shown are the two graphs where the arc from \gls{lnl} III to IV was removed. Their respective model evidence fell below $-1528$ and the two graphs would appear far left in the figure. In the bottom left corner, we provide a visual reference in analogy to \autoref{table:bayes_factor}: E.g., any difference in the model evidence shorter than the first of the three rulers indicates that the improvement is ``barely worth a mention''. \label{fig:ranking}}
\end{figure*}

@@

\begin{deluxetable}{lr}
    \tablecaption{Model comparison results from the base graph and the extended graph we chose as the ``winning'' model. For both \glspl{dag} we show the log-evidence, computed via thermodynamic integration, the negative one half of the \gls{bic}, as well as the maximum log-likelihood that was encountered during the final \gls{mcmc} sampling round. \label{table:evidence}}
    \tablehead{
        \textbf{graph}  & \textbf{log-evidence}
    }
    \startdata
    \variable{output/metrics_table.tex}
    \enddata
\end{deluxetable}

The most likely model parameters for the winning graph, corresponding to the mean of the marginals of the sampled posterior distributions, are tabulated in \autoref{table:params}. We have fixed $p_\text{early}=0.3$ for early T-category tumors (i.e., T0, T1, and T2), and $t_\text{max}=10$ time steps. The result that $t_{\text{II} \rightarrow \text{III}}$ and $t_{\text{III} \rightarrow \text{IV}}$ are relatively large compared to $b_\text{III}$ and $b_\text{IV}$ reflects the observation that skip metastases in levels III and IV without involvement of the upstream level are rare. Since level II's parent node (level I) is rarely involved, $b_\text{II}$ can approximately be related to the prevalence of involvement in level II. The probability for no involvement of level II when the patient is diagnosed after $t$ time steps is $\left(1-b_\text{II}\right)^t$. The prevalence of level II involvement for advanced T-category patients is thus
%
\begin{equation}
    \begin{aligned}
        \text{prev}_\text{late}^\text{II} &= 1-\sum_{t=0}^{10} \left(1-b_\text{II}\right)^t \cdot p_\text{late}^t \, (1-p_\text{late})^{(10-t)}\binom{10}{t} \\
        &\approx 79 \%
    \end{aligned}
\end{equation}
%
which agrees with the second panel from the top in \autoref{fig:bg_prevalences}. 
The large value for the parameter $t_{\text{I} \rightarrow \text{II}}$ reflects the observation that in almost all patients with level I involvement, level II is also involved. The large uncertainty in $t_{\text{I} \rightarrow \text{II}}$ is related to the fact that level I involvement is rare compared to level II.

\begin{deluxetable}{lrr}
    \tablecaption{Mean and standard deviation of parameters sampled for the winning graph in percent. \label{table:params}}
    \tablehead{
        \textbf{parameter} & \textbf{mean} & \textbf{std. dev.}
    }
    \startdata
    \variable{output/means_table.tex}
    \enddata
\end{deluxetable}

@@

\begin{figure}
    \script{wg_any_risks.py}
    \begin{centering}
        \includegraphics[width=\linewidth]{figures/wg_any_risks.png}
        \caption{Shown are the histograms over the predicted risk for involvement in any of the \glspl{lnl} I, IV, V, or VII. The risk is plotted given a clinical N0 diagnosis (green), macroscopically detected metastases in \gls{lnl} II (blue), and lastly given visible involvement in both \gls{lnl} II and III (orange). The top row shows these risks for early T-category diagnoses and the bottom row for advanced T-category. \label{fig:wg_any_risks}}
    \end{centering}
\end{figure}

@@

@@

\begin{figure}
    \script{wg_sens_spec_risks.py}
    \begin{centering}
        \includegraphics[width=\linewidth]{figures/wg_sens_spec_risks.png}
        \caption{Dependence of risk of occult disease on sensitivity and specificity. The figure shows the risk of involvement in level IV, given the mean of the parameter samples, drawn during sampling, in patients with clinical involvement of levels II and III. \label{fig:wg_sens_spec_risks}}
    \end{centering}
\end{figure}

>>>>

\begin{figure}
    \script{thermo_int.py}
    \centering
    \includegraphics[width=\linewidth]{figures/thermo_int.png}
    \caption{The expected log-accuracy under the power posterior plotted against the value of the inverse temperature $\beta$. These values were computed during a \gls{ti} run for the loosing graph model (red line), as well as the winning graph model (green line). Note that the scaling of the x-axis is chosen such that the 64 points on the temperature ladder appear evenly spaced. This is to stress how the log-accuracy develops in the range from $\beta=0$ to around $\beta=0.1$. The area under these two curves yields an approximation of the log-evidence of the respective model. It is barely visible that due to the simplicity of the loosing graph model its accuracy increases earlier and faster that the winning graph model's. However, already before $\beta=0.031$ the model based on the winning graph outperforms the loosing graph model. Ultimately, the winning graph's log-evidence is better than the loosing graph model's by a value of around \variable{output/evidence_diff.txt}(see \autoref{table:evidence}). \label{fig:thermo_int}}
\end{figure}

@@

@@

\begin{deluxetable}{ccccccrrrr}
    \tablecaption{Prevalence of involvement patterns in the multi-centric dataset. An involvement pattern is characterized by the state of the six \glspl{lnl}: A red dot means the \gls{lnl} was reported to be metastatic, a green dot means it was determined to be healthy and a question mark means that the prevalence was marginalized over the state of this \gls{lnl}. \label{table:data_prevalence}}
    \tablehead{
        \multicolumn{6}{c}{\textbf{LNL involvement} } & \multicolumn{4}{c}{\textbf{T-category} } \\
        I & II & III & IV & V & VII & \multicolumn{2}{c}{\textbf{early} } & \multicolumn{2}{c}{\textbf{advanced} }
    }
    \startdata
    \variable{output/data_table.tex}%
    \enddata
\end{deluxetable}

\begin{figure*}[h]
    \script{bg_core_prevs.py}
    \begin{centering}
        \includegraphics[width=\textwidth]{figures/bg_core_prevs.png}
        \caption{Prevalence of involvement as predicted by the base graph model for different scenarios involving the most commonly metastatic \glspl{lnl} II, III and IV (shaded histograms). The model's predictions are compared to Beta posteriors over the prevalence based on the frequency of the same scenarios and a uniform prior (slid lines). The top panel shows some selected scenarios with early T-category tumors and the bottom panel with advanced T-category. \label{fig:bg_prevalences}}
    \end{centering}
\end{figure*}

@@

@@

The dataset containing the consensus decision for the involvement of each level in every patient was then used for model parameter learning. We assumed that it represents an observation of the true hidden state $\boldsymbol{\xi}_k$. The frequencies of some of the most important combinations of involved lymph node levels are listed in \autoref{table:data_prevalence}.
\clearpage

\begin{figure*}[!ht]
    \centering
    \includegraphics[width=\linewidth]{figures/ranking.pdf}
    \caption{Visual ranking of the investigated graphs w.r.t. their model evidence, computed via thermodynamic integration. Not shown are the two graphs where the arc from \gls{lnl} III to IV was removed. Their respective model evidence fell below $-1528$ and the two graphs would appear far left in the figure. In the bottom left corner, we provide a visual reference in analogy to \autoref{table:bayes_factor}: E.g., any difference in the model evidence shorter than the first of the three rulers indicates that the improvement is ``barely worth a mention''. \label{fig:ranking}}
\end{figure*}

@@

@@

\begin{deluxetable}{lr}[t]
    \tablecaption{Model comparison results from the base graph and the extended graph we chose as the ``winning'' model. For both \glspl{dag} we show the log-evidence, computed via thermodynamic integration, the negative one half of the \gls{bic}, as well as the maximum log-likelihood that was encountered during the final \gls{mcmc} sampling round. \label{table:evidence}}
    \tablehead{
        \textbf{graph}  & \textbf{log-evidence}
    }
    \startdata
    \variable{output/metrics_table.tex}%
    \enddata
\end{deluxetable}

@@

\begin{deluxetable}{lrr}[t]
    \tablecaption{Mean and standard deviation of parameters sampled for the winning graph in percent. \label{table:params}}
    \tablehead{
        \textbf{parameter} & \textbf{mean} & \textbf{std. dev.}
    }
    \startdata
    \variable{output/means_table.tex}%
    \enddata
\end{deluxetable}

@@

The most likely model parameters for the winning graph, corresponding to the mean of the marginals of the sampled posterior distributions, are tabulated in \autoref{table:params}. We have fixed $p_\text{early}=0.3$ for early T-category tumors (i.e., T0, T1, and T2), and $t_\text{max}=10$ time steps. The result that $t_{\text{II} \rightarrow \text{III}}$ and $t_{\text{III} \rightarrow \text{IV}}$ are relatively large compared to $b_\text{III}$ and $b_\text{IV}$ reflects the observation that skip metastases in levels III and IV without involvement of the upstream level are rare. Since level II's parent node (level I) is rarely involved, $b_\text{II}$ can approximately be related to the prevalence of involvement in level II. The probability for no involvement of level II when the patient is diagnosed after $t$ time steps is $\left(1-b_\text{II}\right)^t$. The prevalence of level II involvement for advanced T-category patients is thus

\begin{equation}
    \begin{aligned}
        \text{prev}_\text{late}^\text{II} &= 1-\sum_{t=0}^{10} \left(1-b_\text{II}\right)^t \cdot p_\text{late}^t \, (1-p_\text{late})^{(10-t)}\binom{10}{t} \\
        &\approx 79 \%
    \end{aligned}
\end{equation}
%
which agrees with the second panel from the top in \autoref{fig:bg_prevalences}. 
The large value for the parameter $t_{\text{I} \rightarrow \text{II}}$ reflects the observation that in almost all patients with level I involvement, level II is also involved. The large uncertainty in $t_{\text{I} \rightarrow \text{II}}$ is related to the fact that level I involvement is rare compared to level II.

@@

\begin{figure}[b]
    \script{wg_any_risks.py}
    \begin{centering}
        \includegraphics[width=\linewidth]{figures/wg_any_risks.png}
        \caption{Shown are the histograms over the predicted risk for involvement in any of the \glspl{lnl} I, IV, V, or VII. The risk is plotted given a clinical N0 diagnosis (green), macroscopically detected metastases in \gls{lnl} II (blue), and lastly given visible involvement in both \gls{lnl} II and III (orange). The top row shows these risks for early T-category diagnoses and the bottom row for advanced T-category. \label{fig:wg_any_risks}}
    \end{centering}
\end{figure}

@@

\begin{figure}[t]
    \script{wg_sens_spec_risks.py}
    \begin{centering}
        \includegraphics[width=\linewidth]{figures/wg_sens_spec_risks.png}
        \caption{Dependence of risk of occult disease on sensitivity and specificity. The figure shows the risk of involvement in level IV, given the mean of the parameter samples, drawn during sampling, in patients with clinical involvement of levels II and III. \label{fig:wg_sens_spec_risks}}
    \end{centering}
\end{figure}

@@
